%!TEX output_directory = target
% Currently this document is written in German
% !TeX spellcheck = de_DE

%Ensure that all odl school LaTeX habits are remarked
\RequirePackage[l2tabu, orthodox]{nag}
%Neue deutsche Trennmuster
%Siehe http://www.ctan.org/pkg/dehyph-exptl und http://projekte.dante.de/Trennmuster/WebHome
%Nur für pdflatex, nicht für lualatex
\RequirePackage[ngerman=ngerman-x-latest]{hyphsubst}
% \documentclass{lni}
% in Englisch stattdessen:
\documentclass[english]{lni}







%% Some packages, no need to be adabted

% enable copy and paste of ligatures (e.g., in "workflow" and umlauts)
\usepackage{cmap}

%Überschrift des Literaturverzeichnisses
%Only in German
\iflnienglish
\else
\renewcommand{\refname}{Literaturverzeichnis}
\fi

%Enable input of umlauts using UTF-8.
\usepackage[utf8]{inputenc}

\usepackage{graphicx}

\usepackage{fancyhdr}

%Kopf- und Fußzeileneinstellungen
\fancypagestyle{lnifirstpage}{
% Löscht alle Kopf- und Fußzeileneinstellungen
\fancyhf{}

%Kopfzeile
\fancyhead[RO]{\small Einreichung für: <Konferenz/Workshop>,\linebreak%
Geplant als Veröffentlichung innerhalb der Lecture Notes in Informatics (LNI)
}
%Für den Herausgeber:
%\fancyhead[RO]{\small <Vorname Nachname [et.\ al.\ ]>(Hrsg.): <Buchtitel>,\linebreak%
%Lecture Notes in Informatics (LNI), Gesellschaft für Informatik, Bonn <Jahr> \hspace{5pt} \thepage \hspace{0.05cm}}

%Linie unter Kopfzeile
\renewcommand{\headrulewidth}{0.4pt}
}

% Put in the short title (Kurztitle) here
\fancypagestyle{lni}{
\fancyhf{}
%Zu lange Titel müssen von den HerausgeberInnen gekürzt werden, Vorschläge der AutorInnen dazu sind herzlich willkommen.
\fancyhead[RO]{\small Gilbert: Declarative Sparse Linear Algebra on Massively Parallel Dataflow Systems \hspace{5pt}
\thepage \hspace{0.05cm}}
\fancyhead[LE]{\hspace{0.05cm}\small \thepage \hspace{5pt}
%Bis zu drei AutorInnen werden alle angeführt, darüber hinaus wird nur die erste Autorin bzw. der erste Autor angeführt und alle Weiteren mit et al.\ abgekürzt.
%Zu lange AutorInnenlisten müssen von den HerausgeberInnen gekürzt werden.
Till Rohrmann, Sebastian Schelter, Tilmann Rabl and Volker Markl}
\renewcommand{\headrulewidth}{0.4pt}
}

%if lstlistings is used
%better approach: use the minted package - see https://en.wikibooks.org/wiki/LaTeX/Source_Code_Listings#The_minted_package
\usepackage{listings}

\iflnienglish
\usepackage[figurename=Fig., tablename=Tab., font=small]{caption}
\else
\usepackage[figurename=Abb., tablename=Tab., font=small]{caption}
\fi

% Listingname heißt nun List.
\renewcommand{\lstlistingname}{List.}

\usepackage[T1]{fontenc}


%for demonstration purposes only
\usepackage[math]{blindtext}

%tweak \url{...}
\usepackage{url}
\urlstyle{same}
%improve wrapping of URLs - hint by http://tex.stackexchange.com/a/10419/9075
\makeatletter
\g@addto@macro{\UrlBreaks}{\UrlOrds}
\makeatother

%diagonal lines in a table - http://tex.stackexchange.com/questions/17745/diagonal-lines-in-table-cell
%slashbox is not available in texlive (due to licensing) and also gives bad results. Thus, we use diagbox
%\usepackage{diagbox}

\usepackage{booktabs}

%required for pdfcomment later
\usepackage{xcolor}

%for easy quotations: \enquote{text}
%also required by biblatex
\usepackage{csquotes}
\usepackage[
  backend=biber,
  style=LNI
]{biblatex}

% new packages BEFORE hyperref
% See also http://tex.stackexchange.com/questions/1863/which-packages-should-be-loaded-after-hyperref-instead-of-before

%enable hyperref without colors and without bookmarks
\usepackage[
%pdfauthor={},
%pdfsubject={},
%pdftitle={},
%pdfkeywords={},
bookmarks=false,
breaklinks=true,
colorlinks=true,
linkcolor=black,
citecolor=black,
urlcolor=black,
pdfpagelayout=SinglePage,
pdfstartview=Fit
]{hyperref}
%enables correct jumping to figures when referencing
\usepackage[all]{hypcap}

%enable nice comments
\usepackage{pdfcomment}
\newcommand{\commentontext}[2]{\colorbox{yellow!60}{#1}\pdfcomment[color={0.234 0.867 0.211},hoffset=-6pt,voffset=10pt,opacity=0.5]{#2}}
\newcommand{\commentatside}[1]{\pdfcomment[color={0.045 0.278 0.643},icon=Note]{#1}}

%compatibality with TODO package
\newcommand{\todo}[1]{\commentatside{#1}}

% we have to load the user defined packages here in order to not mess up with cleveref
\include{style}

%enable \cref{...} and \Cref{...} instead of \ref: Type of reference included in the link
\iflnienglish
\usepackage[capitalise,nameinlink]{cleveref}
%Nice formats for \cref
\crefname{section}{Sect.}{Sect.}
\Crefname{section}{Section}{Sections}
\crefname{figure}{Fig.}{Fig.}
\Crefname{figure}{Figure}{Figures}
\else
\usepackage[ngerman,capitalise,nameinlink]{cleveref}
\fi

%introduce \powerset - hint by http://matheplanet.com/matheplanet/nuke/html/viewtopic.php?topic=136492&post_id=997377
\DeclareFontFamily{U}{MnSymbolC}{}
\DeclareSymbolFont{MnSyC}{U}{MnSymbolC}{m}{n}
\DeclareFontShape{U}{MnSymbolC}{m}{n}{
    <-6>  MnSymbolC5
   <6-7>  MnSymbolC6
   <7-8>  MnSymbolC7
   <8-9>  MnSymbolC8
   <9-10> MnSymbolC9
  <10-12> MnSymbolC10
  <12->   MnSymbolC12%
}{}
\DeclareMathSymbol{\powerset}{\mathord}{MnSyC}{180}

%improve float placement
%source: http://people.cs.uu.nl/piet/floats/node1.html
%see also: http://tex.stackexchange.com/a/35130/9075
\renewcommand{\textfraction}{0.05}
\renewcommand{\topfraction}{0.95}
\renewcommand{\bottomfraction}{0.95}
\renewcommand{\floatpagefraction}{0.35}
\setcounter{totalnumber}{5}

%enable margin kerning
\usepackage{microtype}

% correct bad hyphenation here
\hyphenation{net-works semi-conduc-tor}

%%% Adapt to your needs from here

%Beginn der Seitenzählung für diesen Beitrag
\setcounter{page}{1}

\author{
Till Rohrmann\footnote{Apache Software Foundation, trohrmann@apache.org} \, %
Sebastian Schelter\footnote{Technische Universität Berlin, sebastian.schelter@tu-berlin.de} \, %
Tilmann Rabl\footnote{Technische Universität Berlin / DFKI, rabl@tu-berlin.de} \, %
Volker Markl\footnote{Technische Universität Berlin / DFKI, volker.markl@tu-berlin.de} \, %
}

\title{Gilbert: Declarative Sparse Linear Algebra\\ on Massively Parallel Dataflow Systems}

\bibliography{paper}

\begin{document}
\maketitle

%hint by http://tex.stackexchange.com/a/30229/9075 and http://tex.stackexchange.com/a/247652/9075
\thispagestyle{lnifirstpage}
\pagestyle{lni}

%Auf Anzahl der AutorInnen setzen, damit die weitere Nummerierung der Fußnoten passt
%Dieser Befehl \verb|\setcounter{footnote}{2}| legt in dem Fall fest, dass 2 Fußnotennummern bereits für die AutorInnen verbraucht wurden, damit die darauf folgenden Fußnoten mit der richtigen Nummerierung (ab 3) fortfahren. Dieser Wert muss an die jeweilige Zahl an AutorInnen bzw. bereits verbrauchte Fußnoten angepasst werden, sofern im weiteren Verlauf Fußnoten verwendet werden.
\setcounter{footnote}{2}

\begin{abstract}
In recent years, the generated and collected data is increasing at an almost exponential rate. 
At the same time, the data's value has been identified in terms of insights that can be provided. 
However, retrieving the value requires powerful analysis tools, since valuable insights are buried deep in large amounts of noise. 
Unfortunately, analytic capacities did not scale well with the growing data. 
Many existing tools run only on a single computer and are limited in terms of data size by its memory. 
A very promising solution to deal with large-scale data is scaling systems and exploiting parallelism.

In this paper, we propose Gilbert, a distributed sparse linear algebra system, to decrease the imminent lack of analytic capacities. 
Gilbert offers a MATLAB\textsuperscript{\textregistered}-like programming language for linear algebra programs, which are automatically executed in parallel. 
Transparent parallelization is achieved by compiling the linear algebra operations first into an intermediate representation. 
This language-independent form enables high-level algebraic optimizations.
Different optimization strategies are evaluated and the best one is chosen by a cost-based optimizer. 
The optimized result is then transformed into a suitable format for parallel execution. 
Gilbert generates execution plans for Apache Spark\textsuperscript{\textregistered} and Apache Flink\textsuperscript{\textregistered}, two massively parallel dataflow systems. 
Distributed matrices are represented by square blocks to guarantee a well-balanced trade-off between data parallelism and data granularity.

An exhaustive evaluation indicates that Gilbert is able to process varying amounts of data exceeding the memory of a single computer on clusters of different sizes.
Two well known machine learning (ML) algorithms, namely PageRank and Gaussian non-negative matrix factorization (GNMF), are implemented with Gilbert. 
The performance of these algorithms is compared to optimized implementations based on Spark and Flink. 
Even though Gilbert is not as fast as the optimized algorithms, it simplifies the development process significantly due to its high-level programming abstraction.
\end{abstract}

\begin{keywords}
Dataflow Optimization, Linear Algebra, Distributed Dataflow Systems
\end{keywords}

\input{1_introduction}
%\input{2_overview}
\input{3_gilbert_features}
\input{4_gilbert_runtime}
\input{5_evaluation}
\input{6_related_work}
\input{7_conclusion}

%No \bibliography{paper} allowed here - biblatex expects it at the preamble
%Run `biber paper` to generate the bibliography
\AtNextBibliography{\footnotesize}
\printbibliography

\end{document}